The Hypothesis of Testing: Paradoxes arising out of reported coronavirus case-counts

### 2020-05-21 write-up/scifile.tex
science-paper

\documentclass[aoas]{amsart}

\RequirePackage[OT1]{fontenc}

\usepackage[foot]{amsaddr}
% \usepackage{biblatex}
\usepackage{natbib}
%\bibliographystyle{apalike}

\usepackage{tchdr}
\boldshortcuts

\usepackage{graphicx}
\usepackage{wrapfig}
\usepackage{setspace}
\DeclareGraphicsExtensions{.eps, .ps}
\usepackage{hyperref}
\usepackage{amsmath, amsthm, amsfonts}
% \usepackage[margin=1.5in]{geometry}

% \numberwithin{equation}{section}
% \theoremstyle{plain}
% \newtheorem{theorem}{Theorem}[section]
% \newtheorem{example}{Example}
% \newtheorem{remark}{Remark}
% \newtheorem{corollary}[theorem]{Corollary}
% \newtheorem{criterion}[theorem]{Criterion}
% \newtheorem{definition}[theorem]{Definition}
% \newtheorem{exercise}[theorem]{Exercise}
% \newtheorem{lemma}[theorem]{Lemma}
% \newtheorem{proposition}[theorem]{Proposition}

\def\pr{\text{pr}}
\def\sgn{\text{sgn}}
\def\I{\bf I}

\begin{document}
\title[The Hypothesis of Testing]{The Hypothesis of Testing: Paradoxes arising out of reported coronavirus case-counts} %

\author{Walter Dempsey}
\address{Department of Biostatistics, University of Michigan, Ann Arbor, MI 48109}
% \email{wdem@umich.edu}

\begin{abstract}
Many statisticians, epidemiologists, economists and data scientists have registered serious reservations regarding the reported coronavirus case-counts. Limited testing capacity across the country has been widely identified as a key driver of suppressed coronavirus case-counts.  The calls to increase testing capacity are well-justified as they become a more frequent point of discussion in the public sphere.  While expanded testing is a laudable goal, selection bias will impact estimates of disease prevalence and the effective reproduction number until the entire population is sampled.  Moreover, tests are imperfect as false positive/negative rates interact in complex ways with selection bias.  In this paper, we attempt to clarify this interaction.  Through simple calculations, we demonstrate pitfalls and paradoxes that can arise when considering case-count data in the presence of selection bias and measurement error. The discussion guides several suggestions on how to improve current case-count reporting.
\end{abstract}

@@

\section{Introduction}
The World Health Organization has declared the coronavirus disease 2019 (COVID-19) a public health emergency.  As of April 27th, 2020, a total of 2,993,000 cases have been confirmed worldwide.  The New York Times reports at least 965,214 people across the United States have tested positive for the virus, and at least 49,465 patients with the virus have died.  Aggressive policies have been put in place across the US with at least 50\% of the US population officially urged as of late April to stay home via state-wide executive actions.

Despite these steps, the data landscape for understanding COVID-19 remains limited.  Public databases maintained by \href{https://bit.ly/2UqFSuA}{Johns Hopkins University}~\citep{JHU_Lancet} and the \href{https://bit.ly/2vUHfrK}{New York Times}~\cite{NYT} provide incoming county-level information of confirmed cases and deaths.  Statisticians, epidemiologists, economists, and data scientists have used this granular data to forecast COVID-19 case-counts, deaths, and hospitalizations~\citep{Song2020,Ray2020,2020.IHME}.  In many cases, variations on the SIR models have been used to draw inferences about infection rates and intensity.  Parametric inference often uses observed case-count and/or observed COVID-19 related deaths to infer latent trajectories of the pandemic.

The goal of this paper is to express reservations at the use of case-counts as a proxy for disease prevalence and intensity as well as its use as direct input into estimation of standard epidemiological models for inference and forecasting.  The reason is straightforward: current models do not account simultaneously for selection bias and measurement error. Selection bias enters due to differences in testing strategies across countries and states.  In the US, for example, limited testing capacity has caused local and state health departments to focus on testing only high-risk populations.  Moreover, testing requires the individual to self-select into asking for a test and then receiving approval by their doctor.  Demands for increased testing capacity, while laudable, ignore the issues of self-selection and measurement error. While increasing testing capacity increases data quantity, there is no guarantee for increased data quality.  We will show that aggressive pushes for ramped up testing capacity that are tied to decreases in data quality may have a mitigated impact  in our ability to estimate quantities of interest such as prevalence and effective reproduction number.

In this paper, we will demonstrate the complex interaction between these two fundamental concepts.  In classical statistics, measurement error is often associated with parameter attenuation; we will show, however, that the combination with self-selection can cause bias in parameter estimates to change sign and increase/decrease in magnitude.  Moreover, the bias may depend on the sampling fraction, prevalence, and population size.  Without a complete understanding of these interactions, we are destined to misinterpret case-counts and come to erroneous conclusions.   We will demonstrate why there is a strong need for more precision and care in presenting COVID-19 analyses to the broader research and non-academic communities.

% \subsection{Related work}

% There has been an abundance..

% However, most complain about testing \emph{capacity} (cite Nate Silver), there is clearly issues of data quality.

% measurement error discussions.
% Selection bias . Proponents of this critique suggest well-designed studies.

% Stanford study was Ionides.  They however, forgot about Critique \#2: measurement

\subsection{Outline}

This article discusses the relationship between three statistical concepts: selection bias, measurement error, and the too often forgotten population size. We clarify mathematically why the situation is much more complex than it first appears.  Through simple mathematical arguments, we clarify five important issues related to COVID-19 case-count analysis.  First, we show that unadjusted prevalence rates are (unsurprisingly) biased when tests are imperfect. What is surprising is that the direction and magnitude of the bias can vary substantially and interacts with the sampling fraction.  Next we show that data quality for adjusted prevalence rates includes an additional term in the error decomposition that captures the interplay among measurement error, selection bias, and prevalence.  Third, we show that daily trends in rate of positive tests is equally problematic even if testing rates and selection protocols are held constant.  We show that the rate of change in COVID-19 observed case-counts overestimates the true rate of change prior to the peak time and underestimates it immediately after.  This implies a data scientist analyzing observed rates will be (on average) overly pessimistic in the early stages of the pandemic, and overly optimistic subsequent to the peak times.  We show that estimates of the effective reproduction number may be similarly impacted.  Fourth, we show that cross-country comparisons are difficult at best with population size, sampling fraction, and data quality all interacting to impact null hypothesis testing.  Fifth, we discuss a sensitivity analysis method for estimating data quality and its use in understanding the potential biases in COVID-19 case-count analyses.  We then end the paper with a discussion of the benefits of randomized testing and use of auxiliary information to improve over simple random sampling.

\section{Analysis of case-count data}
\label{section:casecount}

We start with some simple notation.  Let $N$ denote the population size.  For state-level analysis $N$ is the state's total population, while for country-level analysis $N$ is the country's total population.  At a fixed moment in time, let $Y_j$ denote COVID-19 status for the $j$th individual in the population, $j=1,\ldots, N$. Here, like in survey methodology~\citep{Cochran77}, we treat COVID-19 status as a fixed but unknown quantity of interest. For simplicity, we start by ignoring the dynamic nature of the viral outbreak as well as the fact that individuals can recover from the disease and assume either individual $j$ is COVID-19 positive and $Y_j=1$ or is COVID-19 negative and $Y_j=0$. We also let $I_j \in \{0,1\}$ be an indicator that the individual was selected for testing ($I = 1$) or not ($I=0$).

To start, we assume the primary questions of interest are estimates of the overall number of COVID-19 cases and/or disease prevalence. That is, we are interested in either the population total $Y = \sum_{j=1}^N Y_j$ or the population average $\bar Y = Y/N$. Suppose that $n$ tests are performed and we observe the values $y_1, \ldots, y_n \in \{0,1\}$.  Then a natural candidate for prevalence is the proportion of positive tests $\bar y = \frac{1}{n} \sum_{i=1}^n y_i$, and a natural candidate for overall cases is $N \times \bar y$.
Under simple random sampling (SRS) or any other epsem\footnote{equal probability of selection method} design, the above are unbiased estimators of the population-level quantities of interest.  Under SRS, the variance of the estimator can be expressed as $\frac{1}{N-1} \times \frac{1-f}{f} \times \sigma_Y^2$ where $f = n/N$ is the sampling fraction and $\sigma_Y^2 = \frac{1}{N} \sum_{i=1}^N (Y_i - \bar Y)^2 = \bar Y (1- \bar Y)$.

@@

\begin{wrapfigure}{r}{0.5\textwidth}
\centering
\includegraphics[width = 0.4\textwidth]{../methods/figs/US-casecount.png}
\caption{US total tests and total positive tests per day}
\label{fig:ustests}
\vspace{-0.3cm}
\end{wrapfigure}

Second, there have been calls for increased testing.  While important, many conflate increased testing capacity with increased quality of testing.  We will discuss this point further in Section \ref{section:discussion}.  As of May 8th, the US has performed $8,412,095$ tests in total.  Figure~\ref{fig:ustests} shows the trajectory of testing per day and positive cases per day in the US.  The US population is roughly $328$ million, meaning the fraction sampled is $f = 0.026$.  After smoothing, the empirical prevalence is $9.1\%$ on May 8th.  Supposing COVID-19 positive individuals are $2$ times more likely to get tested than those individuals who are COVID-19 negative, the question is ``What is the sample size from a SRS that would yield equivalent MSE in the estimated prevalence?'' Using the above MSE result, Meng showed the effective sample size is equal to $n_{eff} \leq \frac{1}{E_I [ \rho_{I,Y}^2 ]} \frac{f}{1-f}$.  Under these parameters, the effective sample size is $14$.  Recent proposals~\citep{Siddarth2020} have argued in favor of increasing overall testing capacity.  With more tests available, the relative sampling rate may reduce.  Supposing it drops to $1.2$ then the effective sample size will increase to $299$.  See Appendix~\ref{section:effss} for the mathematical details and calculations in additional scenarios.   The above calculations point to a sad state of affairs, saying that the effective sample size even in best case scenarios is not better than a small random sample from the population.

@@

\subsection{Imperfect testing}

Tests are imperfect.  COVID-19 testing is no exception. Here we investigate the interplay between imperfect testing and selection bias.  In discussions of inaccurate testing, the standard assumption is measurement error leads to parameter attenuation.  When paired with selection bias, however, the two sources of error become entangled, and resulting errors can become magnified, muted, or even switch signs.

First we require some additional notation.  Let $P_j$ be an indicator of measurement error, equal to $1$ when we incorrectly measure the outcome and $0$ when we observe the true outcome. We suppose this is a stochastic variable where $\pr(P_j = 1 \mid Y_j = 1) =: FN$ is the false-negative rate and $\pr(P_j = 1 \mid Y_j = 0) =: FP$ is the false-positive rate.  If individual $j$ is selected (i.e., $I_j = 1$) then the observed outcome can be written as $Y_j^{\star} = Y_j(1-P_j) + (1-Y_j) P_j$.  Suppose disease prevalence was estimated as the fraction who tested positive for COVID-19, i.e., $\bar y_n^\star = \frac{\sum_{i=1}^N I_j Y_j^\star}{\sum_{i=1}^N I_j}$.  We can again investigate the error compared to the true prevalence $\bar Y$ in statistical terms:
$$
\bar y_n^\star - \bar Y = \sqrt{\frac{1-f}{f}} \left[ \rho_{I,Y} \times \sigma_Y + \rho_{I,PZ} \times \sigma_{PZ} + \sqrt{\frac{f}{1-f}}  \left( FP - (FP+FN) \bar Y \right) \right] .
$$
where $Z = 1-2Y$; see Appendix~\ref{app:imperfect} for the derivation. The first term in the large brackets represents the perfect testing regime; to this end, we refer to $\rho_{I,Y}$ as the \emph{true data quality}.  The second term represents the interaction between imperfect testing and selection bias. The variable $PZ = 1[Y=0,P=1] - 1[Y=1, P=1]$ is non-zero only when $P=1$ (i.e., the outcome is incorrectly reported); to this end, we refer to $\rho_{I,PZ}$ as the \emph{observed data quality} that accounts for both selection bias and measurement error.  In the appendix, we show the sign of $\rho_{I,PZ}$ is the opposite of the sign of $\rho_{I,Y}$, which implies that the observed data quality adds error in the opposite direction from the true data quality.  Finally, the third term represents the bias due to imperfect testing.

@@

While prior investigations have noted the interaction between measurement error and selection bias~\citep{Beesley2020,Beesley2019,Smeden2019}, the interaction with the sample size relative to the population, i.e., $f$, has largely been ignored.  The above statistical decomposition clarifies the importance of this quantity~$f$.  In particular, note that the statistical error also includes a bias term due to measurement error and this term increases as the sampled fraction $f$ increases. Therefore, how the first two terms interact with the final term depends on the fraction of the population sampled.  This interaction is complex, but implies that whether the estimate $\bar y^\star_n$ is an overestimate or underestimate is a complicated question due to the relation amongst these three pieces.

@@

The attentive data analyst will recognize the estimator $\bar y_n$ is biased even for simple random samples and, if sensitivity and specificity were known a priori, may suggest the alternative estimator $\tilde y_n = \bar y_n + (1-\bar y_n) FP + FN \bar y_n$ which is unbiased under simple random sampling. We again wish to express the error $\tilde  y_n - \bar Y$ in statistical terms. In the appendix, we show that the error now can be expressed as
\begin{equation}
\label{eq:error}
\rho_{I,Y} \times \sqrt{\frac{1-f}{f}} \times \sigma_{Y}
\times \underbrace{\left[ 1 + FP + FN - \Delta \times \frac{\bar Y}{1-\bar Y} \times \frac{FP(1-\bar Y) + FN \cdot \bar Y}{f_0 (1-\bar Y) + f_1 \bar Y} \right]}_{D_M}.
\end{equation}
The first term is the same as before but increased by $(1 + FP + FN)$ to account for the additional uncertainty due to measurement error.  The second term is the interaction between selection bias and measurement error.  Figure~\ref{fig:heatmap} presents the measurement error data quality adjustment as a function of the relative frequency and odds ratio.  We see that the adjustment can be both positive and negative as well as a range of magnitudes.

@@

\begin{wrapfigure}{r}{0.5\textwidth}
\centering
\includegraphics[width = 0.4\textwidth]{../methods/figs/mem_heatmap_article.png}
\caption{Measurement error data quality adjustment: relative frequency $f_1/f_0$ (x-axis) against odds ratio (y-axis) for $FP=0.005$ and $FN=0.172$. Color scaled so blue = $-10$, white = $0$, and red = $10$.}
\label{fig:heatmap}
\vspace{-0.3cm}
\end{wrapfigure}

@@

\subsection{Regrettable rates: complex biases resulting from self-selection}
\label{section:rates}

@@

Consider the state of New York.  Figure~\ref{fig:ny-covid-test} shows the number of tests and positive tests per day in New York state.  Figure~\ref{fig:ny-covid-frac} presents the fraction of tests that were positive per day in New York state.  For cancellation, we would need the expected data qualities to be uncorrelated and equal. Testing appears relatively stable over a week which suggests assuming $\Delta_t$ (i.e., differential in the selection rate) varies slowly over time may be a plausible assumption.  Note, however, that constant $\Delta_t$ does not imply constant data quality (see Appendix \ref{section:effss} for details).  Figure~\ref{fig:ny-covid-test} shows the number of tests has remained relatively constant in the last few weeks, specifically relative to the total NY population, assuming $f_t$ also varies slowly in time is reasonable. Finally, the measurement error adjustment is likely to be constant as no new testing procedures have been reported.  This would not be true in states like Michigan, where testing criteria have been expanding every few weeks.  The final component is an adjustment to this differential error based on the statistical error at the first time point.

Figure \ref{fig:ratio-bias} shows a trajectory of the true ratio and the potential biased estimators under an SIR model for the epidemic dynamics, with state evolution given by the following ODE
$$
\begin{aligned}
\frac{\partial S_t}{\partial t} &= - \beta S_t I_t N^{-1}; \quad
\frac{\partial I_t}{\partial t} = \beta S_t I_t N^{-1} - \gamma I_t; \quad
\frac{\partial R_t}{\partial t} = - \gamma I_t \\
\end{aligned}
$$
where $S_t$ is the number of susceptible individuals at time $t$, $I_t$ is the number of infected individuals at time $t$, $N$ is the total population size, and $R_t$ is the number of removed (recovered or deceased) individuals at time $t$.   Figure~\ref{fig:fracpop} presents the fraction of population newly infected individuals at each time step for $\beta = 1.4$ and $\gamma = 0.2$ (black line).  In each case, the rate of change in prevalence is likely to be  overestimated prior to the peak and then underestimated afterwards.  From a decision making perspective, such biases have potential impact.  First, the over-estimation may give lawmakers and governors more leverage in proposing aggressive actions that reduce prevalence.  Of course, the analysis supports the argument that the estimates based on available data are overly pessimistic. What both sides miss is that the direction of bias is non-constant over time.  After we hit the peak, estimates will likely under-estimate prevalence.  Such bias puts pressure on lawmakers and governors to prematurely relax social distancing measures.  Moreover, the question is ``Can we trust the observed data to let us know if we have reached the peak?''  It appears that the peak time is the easiest to estimate in terms of having minimal error using availability data.  Note, however, that the standard errors on such estimates will be incredibly wide due to the low effective sample size as discussed in Section~\ref{section:casecount}.

\subsection{Estimation of effective reproduction number}

@@

Many well-respected epidemiologists argue that tracking the effective reproduction number $R_t$ is the only way to manage through the crisis~\citep{Gabriel2020}.  We show here how these estimates are also impacted by selection bias and measurement error.  For simplicity, we again assume the estimates are adjusted for known false positive and false negative rates. \cite{Bettencourt2008} show that under a Poisson likelihood, a simple relation between the trajectory of new cases and the effective reproduction number can be derived.  In particular, under an SIR model, the number of case counts on day $t$ is Poisson distributed with rate $K_{t-1} \exp \left( \frac{1}{\gamma} (R_t - 1) \right)$ where $K_{t-1} = Y_{t-1}-Y_{t-2}$ is the number of new cases on day $t-1$ and $\gamma$ is the serial interval which is approximately $7$ days for COVID-19~\citep{Sanche2020}.  Using this, a simple estimator of the effective reproduction number at time $t$ can be given by

@@

Randomization of treatment assignment in clinical trials negates unobserved confounders.  This, however, does not negate effect modifiers.  Therefore, for marginal treatment effects to be interpretable, there must be a well-defined population.  Most often, our main interest is in causal effects on the population of COVID-19 positive.  Randomization within the clinical trial design yields internal validity, but we also need external validity to estimate the correct marginal effect of interest.

\section{Cross-population comparisons}

@@

\subsection{Prevalence comparisons}

@@

\begin{figure}
\centering
\begin{subfigure}{.5\textwidth}
  \centering
  \includegraphics[width=.9\linewidth]{../methods/figs/NY_casecount.png}
  \caption{NY new cases per day}
  \label{fig:ny-covid-test}
\end{subfigure}%
\begin{subfigure}{.5\textwidth}
  \centering
  \includegraphics[width=.9\linewidth]{../methods/figs/NY_fracpos.png}
  \caption{NY fraction of new tests positive per day}
  \label{fig:ny-covid-frac}
\end{subfigure}
\caption{New York COVID-19 reported data }
\label{fig:ny-covid}
\end{figure}

\section{Stratified sampling: improving precision in low-prevalence environments}

@@

\subsection{Ratio estimator}
\label{app:ratio}

>>>>

% Use only LaTeX2e, calling the article.cls class and 12-point type.

\documentclass[12pt]{article}

% Users of the {thebibliography} environment or BibTeX should use the
% scicite.sty package, downloadable from *Science* at
% http://www.sciencemag.org/authors/preparing-manuscripts-using-latex
% This package should properly format in-text
% reference calls and reference-list numbers.

\usepackage{scicite}

\usepackage{times}
\usepackage{amsmath,amsthm,amsfonts}

% The preamble here sets up a lot of new/revised commands and
% environments.  It's annoying, but please do *not* try to strip these
% out into a separate .sty file (which could lead to the loss of some
% information when we convert the file to other formats).  Instead, keep
% them in the preamble of your main LaTeX source file.

% The following parameters seem to provide a reasonable page setup.

\topmargin 0.0cm
\oddsidemargin 0.2cm
\textwidth 16cm
\textheight 21cm
\footskip 1.0cm

%The next command sets up an environment for the abstract to your paper.

\newenvironment{sciabstract}{%
\begin{quote} \bf}
{\end{quote}}

\def\I{{\bf I}}
\def\bary{{\bar y}}
\def\pr{\text{pr}}
\def\sgn{\text{sgn}}

% Include your paper's title here

\title{The Hypothesis of Testing: Paradoxes arising out of reported coronavirus case-counts}

% Place the author information here.  Please hand-code the contact
% information and notecalls; do *not* use \footnote commands.  Let the
% author contact information appear immediately below the author names
% as shown.  We would also prefer that you don't change the type-size
% settings shown here.

\author
{Walter Dempsey$^{1\ast}$\\
\\
\normalsize{$^{1}$Department of Biostatistics, University of Michigan,}\\
\normalsize{1415 Washington Heights, Ann Arbor, MI 48109, USA}\\
\\
\normalsize{$^\ast$To whom correspondence should be addressed; E-mail:  wdem@umich.edu.}
}

% Include the date command, but leave its argument blank.

\date{}

%%%%%%%%%%%%%%%%% END OF PREAMBLE %%%%%%%%%%%%%%%%

\begin{document}

% Double-space the manuscript.

\baselineskip24pt

% Make the title.

@@

% Place your abstract within the special {sciabstract} environment.

\begin{sciabstract}
  Many statisticians, epidemiologists, economists and data scientists have registered serious reservations regarding the reported coronavirus case-counts. Limited testing capacity across the country has been widely identified as a key driver of suppressed coronavirus case-counts.  The calls to increase testing capacity are well-justified as they become a more frequent point of discussion in the public sphere.  While expanded testing is a laudable goal, selection bias will impact estimates of disease prevalence and the effective reproduction number until the entire population is sampled.  Moreover, tests are imperfect as false positive/negative rates interact in complex ways with selection bias.  In this paper, we attempt to clarify this interaction.  Through simple calculations, we demonstrate pitfalls and paradoxes that can arise when considering case-count data in the presence of selection bias and measurement error. The discussion guides several suggestions on how to improve current case-count reporting.
\end{sciabstract}

% In setting up this template for *Science* papers, we've used both
% the \section* command and the \paragraph* command for topical
% divisions.  Which you use will of course depend on the type of paper
% you're writing.  Review Articles tend to have displayed headings, for
% which \section* is more appropriate; Research Articles, when they have
% formal topical divisions at all, tend to signal them with bold text
% that runs into the paragraph, for which \paragraph* is the right
% choice.  Either way, use the asterisk (*) modifier, as shown, to
% suppress numbering.

\section*{Introduction}

The World Health Organization has declared the coronavirus disease 2019 (COVID-19) a public health emergency.  As of May 21st, 2020, over 5 million cases have been confirmed worldwide with 1.5 million confirmed cases and over 90 thousand COVID-19 deaths across the United States. This pandemic
%has ripped through our communities and
has become the focal point of every day life; yet the data landscape for understanding COVID-19 remains limited.  Public databases maintained by Johns Hopkins University~\cite{JHU_Lancet} and the New York Times~\cite{NYT} provide incoming county-level information of confirmed cases and deaths.  Statisticians, epidemiologists, economists, and data scientists have attempted to use this granular data to forecast COVID-19 case-counts, deaths, and hospitalizations~\cite{Song2020,Ray2020,2020.IHME}.

The goal of this paper is to express reservations at the use of case-counts as a proxy for disease prevalence and in estimation of standard epidemiological models for inference and forecasting.  The reason is straightforward: current models do not account simultaneously for selection bias and measurement error. In the US, for example, limited testing capacity has caused health departments to test only high-risk individuals.  Moreover, testing requires individuals to request a test and receive approval from their doctor.  Demands for increased testing capacity, while laudable, do not guarantee increased data quality.  Aggressive measures to increase testing capacity that decrease data quality may have a mitigated impact in our ability to estimate quantities of interest such as prevalence and effective reproduction number. We will demonstrate why there is a strong need for more precision and care in presenting COVID-19 analyses to the broader research and non-academic communities.

\section*{Analysis of imperfect case-count data}
\label{section:casecount}

Let $N$ denote the population size.  At a fixed time, let $Y_j$ denote COVID-19 status for the $j$th individual in the population, $j=1,\ldots, N$. Here, as in survey methodology~\cite{Cochran77}, we treat COVID-19 status as a fixed but unknown quantity of interest. For simplicity, we start by ignoring the dynamic nature of the outbreak and the fact that individuals can recover.  We also let $I_j \in \{0,1\}$ be an indicator that the individual was selected for testing ($I = 1$) or not ($I=0$).

Here we consider estimates of the number of COVID-19 cases and/or disease prevalence, i.e., the population total $Y = \sum_{j=1}^N Y_j$ or the population average $\bar Y = Y/N$. Suppose that $n$ tests are performed and we observe the values $y_1, \ldots, y_n \in \{0,1\}$.  Then a natural candidate for prevalence is the proportion of positive tests $\bar y = n^{-1} \sum_{i=1}^n y_i$, and a natural candidate for overall cases is $N \times \bar y$.
Under simple random sampling (SRS), both are unbiased estimators of the  quantities of interest and the variance of the prevalence estimator equals $(N-1)^{-1} \times (1-f)/f \times \sigma_Y^2$ where $f = n/N$ is the sampling fraction and $\sigma_Y^2 = \frac{1}{N} \sum_{i=1}^N (Y_i - \bar Y)^2 = \bar Y (1- \bar Y)$.

In the current context, the true status is not observed. Instead, individuals are tested for COVID-19 status. Tests are imperfect.  COVID-19 testing is no exception. The standard assumption is that measurement error leads to parameter attenuation.  When paired with selection bias, however, the two sources of error become entangled, and resulting errors can become magnified, muted, or even switch signs~\cite{Beesley2020,Beesley2019,Smeden2019}. Let $P_j$ be an indicator of measurement error, equal to $1$ when we incorrectly measure the outcome and $0$ when we observe the true outcome. We suppose this is a stochastic variable where $\pr(P_j = 1 \mid Y_j = 1) =: FN$ is the false-negative rate and $\pr(P_j = 1 \mid Y_j = 0) =: FP$ is the false-positive rate.  If individual $j$ is selected (i.e., $I_j = 1$) then the observed outcome can be written as $Y_j^{\star} = Y_j(1-P_j) + (1-Y_j) P_j$.  Suppose disease prevalence is estimated by $\tilde y_n = \bar y_n + (1-\bar y_n) FP + FN \bar y_n$ which is unbiased under simple random sampling. We wish to express the error $\tilde  y_n - \bar Y$ in statistical terms. In the supplementary materials, we show that the error now can be expressed as
\begin{equation}
\label{eq:error}
\rho_{I,Y} \times \sqrt{\frac{1-f}{f}} \times \sigma_{Y}
\times \underbrace{\left[ 1 + FP + FN - \Delta \times \frac{\bar Y}{1-\bar Y} \times \frac{FP(1-\bar Y) + FN \cdot \bar Y}{f_0 (1-\bar Y) + f_1 \bar Y} \right]}_{D_M}.
\end{equation}

@@

Second, there have been calls for increased testing.  While important, many conflate increased testing capacity with increased quality of testing.  We will discuss this point further in Section \ref{section:discussion}.  As of May 8th, the US has performed $8,412,095$ tests in total.  Figure~1 shows the trajectory of testing per day and positive cases per day in the US.  The US population is roughly $328$ million, meaning the fraction sampled is $f = 0.026$.  After smoothing, the empirical prevalence is $9.1\%$ on May 8th.  Supposing COVID-19 positive individuals are $2$ times more likely to get tested than those individuals who are COVID-19 negative, the question is ``What is the sample size from a SRS that would yield equivalent MSE in the estimated prevalence?'' Using the above MSE result, Meng showed the effective sample size is equal to $n_{eff} \leq \frac{1}{E_I [ \rho_{I,Y}^2 ]} \frac{f}{1-f}$.  Under these parameters, the effective sample size is $14$.  Recent proposals~\cite{Siddarth2020} have argued in favor of increasing overall testing capacity.  With more tests available, the relative sampling rate may reduce.  Supposing it drops to $1.2$ then the effective sample size will increase to $299$.  See Appendix~\ref{section:effss} for the mathematical details and calculations in additional scenarios.   The above calculations point to a sad state of affairs, saying that the effective sample size even in best case scenarios is not better than a small random sample from the population.

@@

\subsection*{Imperfect testing}

Let $P_j$ be an indicator of measurement error, equal to $1$ when we incorrectly measure the outcome and $0$ when we observe the true outcome. We suppose this is a stochastic variable where $\pr(P_j = 1 \mid Y_j = 1) =: FN$ is the false-negative rate and $\pr(P_j = 1 \mid Y_j = 0) =: FP$ is the false-positive rate.  If individual $j$ is selected (i.e., $I_j = 1$) then the observed outcome can be written as $Y_j^{\star} = Y_j(1-P_j) + (1-Y_j) P_j$.  Suppose disease prevalence was estimated as the fraction who tested positive for COVID-19, i.e., $\bar y_n^\star = \frac{\sum_{i=1}^N I_j Y_j^\star}{\sum_{i=1}^N I_j}$.  We can again investigate the error compared to the true prevalence $\bar Y$ in statistical terms:
$$
\bar y_n^\star - \bar Y = \sqrt{\frac{1-f}{f}} \left[ \rho_{I,Y} \times \sigma_Y + \rho_{I,PZ} \times \sigma_{PZ} + \sqrt{\frac{f}{1-f}}  \left( FP - (FP+FN) \bar Y \right) \right] .
$$
where $Z = 1-2Y$; see Appendix~\ref{app:imperfect} for the derivation. The first term in the large brackets represents the perfect testing regime; to this end, we refer to $\rho_{I,Y}$ as the \emph{true data quality}.  The second term represents the interaction between imperfect testing and selection bias. The variable $PZ = 1[Y=0,P=1] - 1[Y=1, P=1]$ is non-zero only when $P=1$ (i.e., the outcome is incorrectly reported); to this end, we refer to $\rho_{I,PZ}$ as the \emph{observed data quality} that accounts for both selection bias and measurement error.  In the appendix, we show the sign of $\rho_{I,PZ}$ is the opposite of the sign of $\rho_{I,Y}$, which implies that the observed data quality adds error in the opposite direction from the true data quality.  Finally, the third term represents the bias due to imperfect testing.

@@

While prior investigations have noted the interaction between measurement error and selection bias~\cite{Beesley2020,Beesley2019,Smeden2019}, the interaction with the sample size relative to the population, i.e., $f$, has largely been ignored.  The above statistical decomposition clarifies the importance of this quantity~$f$.  In particular, note that the statistical error also includes a bias term due to measurement error and this term increases as the sampled fraction $f$ increases. Therefore, how the first two terms interact with the final term depends on the fraction of the population sampled.  This interaction is complex, but implies that whether the estimate $\bar y^\star_n$ is an overestimate or underestimate is a complicated question due to the relation amongst these three pieces.

@@

The attentive data analyst will recognize the estimator $\bar y_n$ is biased even for simple random samples and, if sensitivity and specificity were known a priori, may suggest the alternative estimator $\tilde y_n = \bar y_n + (1-\bar y_n) FP + FN \bar y_n$ which is unbiased under simple random sampling. We again wish to express the error $\tilde  y_n - \bar Y$ in statistical terms. In the appendix, we show that the error now can be expressed as
\begin{equation}
\label{eq:error}
\rho_{I,Y} \times \sqrt{\frac{1-f}{f}} \times \sigma_{Y}
\times \underbrace{\left[ 1 + FP + FN - \Delta \times \frac{\bar Y}{1-\bar Y} \times \frac{FP(1-\bar Y) + FN \cdot \bar Y}{f_0 (1-\bar Y) + f_1 \bar Y} \right]}_{D_M}.
\end{equation}
The first term is the same as before but increased by $(1 + FP + FN)$ to account for the additional uncertainty due to measurement error.  The second term is the interaction between selection bias and measurement error.  Figure~2 presents the measurement error data quality adjustment as a function of the relative frequency and odds ratio.  We see that the adjustment can be both positive and negative as well as a range of magnitudes.

@@

@@

\subsection*{Regrettable rates: complex biases resulting from self-selection}
\label{section:rates}

@@

Consider the state of New York.  Figure~5 shows the number of tests and positive tests per day in New York state.  Figure~5b presents the fraction of tests that were positive per day in New York state.  For cancellation, we would need the expected data qualities to be uncorrelated and equal. Testing appears relatively stable over a week which suggests assuming $\Delta_t$ (i.e., differential in the selection rate) varies slowly over time may be a plausible assumption.  Note, however, that constant $\Delta_t$ does not imply constant data quality (see Appendix \ref{section:effss} for details).  Figure~5a shows the number of tests has remained relatively constant in the last few weeks, specifically relative to the total NY population, assuming $f_t$ also varies slowly in time is reasonable. Finally, the measurement error adjustment is likely to be constant as no new testing procedures have been reported.  This would not be true in states like Michigan, where testing criteria have been expanding every few weeks.  The final component is an adjustment to this differential error based on the statistical error at the first time point.

Figure~3a shows a trajectory of the true ratio and the potential biased estimators under an SIR model for the epidemic dynamics, with state evolution given by the following ODE
$$
\begin{aligned}
\frac{\partial S_t}{\partial t} &= - \beta S_t I_t N^{-1}; \quad
\frac{\partial I_t}{\partial t} = \beta S_t I_t N^{-1} - \gamma I_t; \quad
\frac{\partial R_t}{\partial t} = - \gamma I_t \\
\end{aligned}
$$
where $S_t$ is the number of susceptible individuals at time $t$, $I_t$ is the number of infected individuals at time $t$, $N$ is the total population size, and $R_t$ is the number of removed (recovered or deceased) individuals at time $t$.   Figure~4a presents the fraction of population newly infected individuals at each time step for $\beta = 1.4$ and $\gamma = 0.2$ (black line).  In each case, the rate of change in prevalence is likely to be  overestimated prior to the peak and then underestimated afterwards.  From a decision making perspective, such biases have potential impact.  First, the over-estimation may give lawmakers and governors more leverage in proposing aggressive actions that reduce prevalence.  Of course, the analysis supports the argument that the estimates based on available data are overly pessimistic. What both sides miss is that the direction of bias is non-constant over time.  After we hit the peak, estimates will likely under-estimate prevalence.  Such bias puts pressure on lawmakers and governors to prematurely relax social distancing measures.  Moreover, the question is ``Can we trust the observed data to let us know if we have reached the peak?''  It appears that the peak time is the easiest to estimate in terms of having minimal error using availability data.  Note, however, that the standard errors on such estimates will be incredibly wide due to the low effective sample size as discussed in Section~\ref{section:casecount}.

\subsection*{Estimation of effective reproduction number}

@@

Many well-respected epidemiologists argue that tracking the effective reproduction number $R_t$ is the only way to manage through the crisis~\cite{Gabriel2020}.  We show here how these estimates are also impacted by selection bias and measurement error.  For simplicity, we again assume the estimates are adjusted for known false positive and false negative rates. \cite{Bettencourt2008} show that under a Poisson likelihood, a simple relation between the trajectory of new cases and the effective reproduction number can be derived.  In particular, under an SIR model, the number of case counts on day $t$ is Poisson distributed with rate $K_{t-1} \exp \left( \frac{1}{\gamma} (R_t - 1) \right)$ where $K_{t-1} = Y_{t-1}-Y_{t-2}$ is the number of new cases on day $t-1$ and $\gamma$ is the serial interval which is approximately $7$ days for COVID-19~\cite{Sanche2020}.  Using this, a simple estimator of the effective reproduction number at time $t$ can be given by

@@

Randomization of treatment assignment in clinical trials negates unobserved confounders.  This, however, does not negate effect modifiers.  Therefore, for marginal treatment effects to be interpretable, there must be a well-defined population.  Most often, our main interest is in causal effects on the population of COVID-19 positive.  Randomization within the clinical trial design yields internal validity, but we also need external validity to estimate the correct marginal effect of interest~\cite{Keiding2016}.

\section*{Cross-population comparisons}

@@

\subsection*{Prevalence comparisons}

@@

\section*{Discussion}
\label{section:discussion}

\subsection*{Stratified sampling: improving precision in low-prevalence environments}

@@

\subsection*{Ratio estimator}
\label{app:ratio}
